---
title: "Predictive Inference for Cognitive Decline Using Mixed Effects Models"
author: "Daniel Dema"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(spida2)
library(p3d)
library(nlme)
library(car)
library(lattice)
library(latticeExtra)
library(dplyr)
library(glmmTMB)
library(DHARMa)
library(splines)
library(performance)
library(lme4)
library(lmerTest)
```

## Goals

- Predictive inference for cognitive decline in Alzheimer's disease
- Explore the limitations of LMEs on non-normal response data

## Data Overview

- Subject ID
- Visit number
- Sex
- Age at visit
- Years of education
- Socioeconomic status (SES): score from 0-5
- Normalized whole brain volume (nWBV): ranges from 0-1
- Mini-Mental State Examination (MMSE): test score from 0-30
- Clinical Dementia Ratio (CDR): CDR = 0 (non-demented), CDR = 0.5 (very mild Alzheimer's), CDR = 1 (mild Alzheimer's)

***

```{r}
df <- read.csv(
  "D:/Daniel/Documents/MATH6642/final_project/Data/oasis_longitudinal.csv")
dc <- df[!is.na(df$MMSE) & !is.na(df$SES), ]

dc$SES <- 5 - dc$SES

dc_orig <- dc

dc <- dc[order(dc$Subject.ID, dc$Age), ]
```

***

```{r}
hist(dc$EDUC, 
     breaks = 10, 
     main = "Histogram of Education Level", 
     xlab = "Years of Education",
     ylab = "Number of Individuals",
     col = "lightblue",
     border = "white")
```

***

```{r}
barplot(table(dc$SES),
        main = "Barplot of Socioeconomic Status",
        xlab = "SES",
        ylab = "Number of Individuals",
        col = "lightblue",
        border = "white")
```

***

```{r}
barplot(table(dc$M.F), 
        main = "Barplot of Sex", 
        xlab = "Sex", 
        ylab = "Number of Individuals", 
        col = c("pink", "lightblue"),
        border = "white")
```

***

```{r}
initial_visits <- dc[ave(dc$Visit, dc$Subject.ID, FUN = min) == dc$Visit, ]
final_visits <- dc[ave(dc$Visit, dc$Subject.ID, FUN = max) == dc$Visit, ]

#Histogram for initial visits
hist(initial_visits$MMSE, main = "Initial Visit MMSE", breaks = 0:30, xlab = "MMSE",
     ylab = "Number of Individuals", col = "lightblue", border = "white", xlim = c(0, 30))
```

***

```{r}
#Histogram for final visits
hist(final_visits$MMSE, main = "Final Visit MMSE", breaks = 0:30, xlab = "MMSE",
     ylab = "Number of Individuals", col = "lightblue", border = "white", xlim = c(0, 30))
```

## Constructing the LME Model

Starting fixed effects:

- Age
- Years of education (EDUC)
- Socioeconomic status (SES)
- Brain volume (nWBV)
- Sex (M.F)

***

```{r}
fit1 <- lme(MMSE ~ nWBV + Age + M.F + SES + EDUC, 
                  data = dc, 
                  random = ~ 1 + Age | Subject.ID,
                  correlation = corAR1(form = ~ 1 | Subject.ID))
```

***

Random effects possibilities:

- random = ~ 1 | Subject.ID
- random = ~ 1 + Age | Subject.ID
- random = ~ 1 + nWBV

***

random = ~ 1 + nWBV has best AIC performance

```{r}
fit2 <- lme(MMSE ~ nWBV + Age + M.F + SES + EDUC, 
            data = dc, 
            random = ~ 1 + nWBV | Subject.ID,
            correlation = corAR1(form = ~ 1 | Subject.ID),
            control = lmeControl(opt = "optim", maxIter = 200, 
            msMaxIter = 200))
```

***

```{r}
VarCorr(fit2)
```

***

```{r}
summary(fit2)$tTable
```

***

For fixed effects:

Possible 3-way interactions:

- nWBV * Age * SES: SES influences access to healthcare, slowing decay in nWBV as the subject ages.

- nWBV * Age * M.F: Possible differences in nWBV decay by sex as the subject ages.

- nWBV * Age * EDUC: Higher education slows the loss of brain volume as the subject ages.

***

```{r}
fit3 <- lme(MMSE ~ nWBV + Age + M.F + SES + EDUC + nWBV * Age * SES
            + nWBV * Age + nWBV * SES + Age * SES, 
            data = dc, 
            random = ~ 1 + nWBV | Subject.ID,
            correlation = corAR1(form = ~ 1 | Subject.ID),
            control = lmeControl(opt = "optim", maxIter = 200,
                                 msMaxIter = 200))
```

***

```{r}
VarCorr(fit3)
```

***

```{r}
summary(fit3)$tTable
```

***

Let's consider 2-way interactions instead:

- nWBV * Age : Brain volume changes with age.

- nWBV * EDUC: Education acts as a buffer against brain volume loss.

- nWBV * SES : Higher SES allows for better healthcare, improving preservation of brain volume.

- nWBV * M.F : Sex affects brain volume.

- Age * EDUC : Education acts as a buffer against age effects.

- Age * M.F : Aging affects women and men differently.

- Age * SES : Higher SES acts as a buffer against age effects by again allowing for better healthcare.

***

```{r}
fit5 <- lme(MMSE ~ nWBV + Age + SES + nWBV * SES + Age * SES, 
            data = dc, 
            random = ~ 1 + nWBV | Subject.ID,
            correlation = corAR1(form = ~ 1 | Subject.ID),
            control = lmeControl(opt = "optim", maxIter = 200,
            msMaxIter = 200))
```

Lowest AIC so far, all terms statistically significant

***

```{r}
summary(fit5)$tTable
```

***

```{r}
VarCorr(fit5)
```


## Diagnostics for G-Matrix

Try within-subject centering?

```{r}
test <- lme(MMSE ~ nWBV + Age + SES + nWBV * SES + Age * SES, 
            data = dc, 
            random = ~ 1 + dvar(nWBV, Subject.ID) | Subject.ID,
            correlation = corAR1(form = ~ 1 | Subject.ID),
            control = lmeControl(opt = "optim", maxIter = 200,
                                 msMaxIter = 200))

anova(fit5, test)
```

***

```{r}
VarCorr(test)
```

***

```{r}
summary(test)$tTable
```

***

Trade-off:

Higher AIC, loss of interaction term in exchange for stable G-matrix

## Residuals

Reconstruct fit5 as a glmmTMB so we can use DHARMa to plot residuals:

```{r}
fit5_tmb <- glmmTMB(
  MMSE ~ nWBV + Age + SES + nWBV:SES + Age:SES + (1 + nWBV | Subject.ID),
  data = dc,
  REML = TRUE
)

res <- simulateResiduals(fittedModel = fit5_tmb)
```

***

```{r}
plot(res)
```

***

Problems:

- Heteroskedasticity
- KS test: residuals are not uniformly distributed
- Outlier test: model handles outliers poorly

Let's try non-linear terms.

***

```{r}
dc$nWBV_sq <- dc$nWBV^2

fit_nwbv_poly_raw <- glmmTMB(
  MMSE ~ nWBV + nWBV_sq + Age + SES + nWBV:SES + nWBV_sq:SES + Age:SES
  + (1 + nWBV | Subject.ID),
  data = dc,
  REML = TRUE
)

sim_nwbv_poly <- simulateResiduals(fit_nwbv_poly_raw)
```

***

```{r}
plot(sim_nwbv_poly)
```

***

```{r}
AIC(fit_nwbv_poly_raw, fit5_tmb)
```

***

```{r}
dc$nWBV_scaled <- scale(dc$nWBV)[,1]
dc$nWBV_sq_scaled <- dc$nWBV_scaled^2

fit6 <- lmer(MMSE ~ nWBV_scaled + nWBV_sq_scaled + Age + SES + 
             nWBV_scaled * SES + Age * SES + 
             (1 + nWBV_scaled | Subject.ID),
             data = dc)
```

***

```{r}
VarCorr(fit6)
```

***

```{r}
as.data.frame(summary(fit6)$coefficients[, "Pr(>|t|)"])
```

***

What if we try splines?

```{r}
dc$ns_nWBV_scaled <- ns(dc$nWBV_scaled, df = 4)
ns_basis <- ns(dc$nWBV_scaled, df = 4)
ns_df <- as.data.frame(ns_basis)
colnames(ns_df) <- paste0("ns_nWBV_", 1:ncol(ns_df))

dc <- bind_cols(dc, ns_df)

fit_spline_tmb <- glmmTMB(
  MMSE ~ ns_nWBV_1 + ns_nWBV_2 + ns_nWBV_3 + Age + SES
  + nWBV * SES + Age * SES
  + (1 + nWBV | Subject.ID),
  data = dc,
  REML = TRUE
)

res_spline <- simulateResiduals(fit_spline_tmb)
```

***

```{r}
plot(res_spline)
```

***

```{r}
dc$nWBV_scaled <- scale(dc$nWBV)[,1]

fit_spline <- lmer(MMSE ~ ns(nWBV_scaled, df=4) + Age + SES + 
                     nWBV_scaled*SES + Age:SES + 
                   (1 + nWBV_scaled | Subject.ID),
                   data = dc,
                   REML = TRUE)
```

***

```{r}
VarCorr(fit_spline)
```

***

```{r}
as.data.frame(summary(fit_spline)$coefficients[, "Pr(>|t|)"])
```

## Solution for Residuals?

Use a glmmTMB with zero-inflation.

```{r}
dc_c <- dc
dc_c$MMSE <- 30 - dc_c$MMSE
dc_c$nWBV_scaled <- scale(dc_c$nWBV)[,1]
dc_c$Age_scaled <- scale(dc_c$Age)[,1]
dc_c$SES_scaled <- scale(dc_c$SES)[,1]

fit5_tmb <- glmmTMB(
  MMSE ~ nWBV_scaled + Age_scaled + SES_scaled + 
         nWBV_scaled * SES_scaled + Age_scaled * SES_scaled + 
         (1 | Subject.ID),
  data = dc_c,
  ziformula = ~ Age_scaled,
  family = poisson,
  REML = TRUE
)
```

***

```{r}
AIC(fit5, fit5_tmb)
```

***

```{r}
res1 <- simulateResiduals(fit5_tmb)
plot(res1)
```

## Model Interpretation

```{r}
VarCorr(fit_spline)
```

***
```{r}
wald(fit_spline)
```

## Model Comparison

```{r}
AIC(fit5, fit6, fit_spline)
```

***

```{r}
cdr_colors <- c("0" = "lightblue", "0.5" = "violet", "1" = "darkblue")

#nWBV trajectories
nWBV_traj <- xyplot(nWBV ~ Age, data = dc,
       groups = Subject.ID,
       type = "b",
       lwd = 1,
       pch = 16,
       col = cdr_colors[as.character(dc$CDR)],
       xlab = "Age (years)",
       ylab = "nWBV (%)",
       main = "Longitudinal trajectories of nWBV per subject by CDR group",
       key = list(text = list(c("CDR 0", "CDR 0.5", "CDR 1")),
                  points = list(pch = 16, col = c("lightblue", "violet", "darkblue")),
                  columns = 3))
```

***

```{r}
nWBV_traj
```

***

```{r}
#MMSE trajectories
MMSE_traj <- xyplot(MMSE ~ Age, data = dc,
       groups = Subject.ID,
       type = "b",
       lwd = 1,
       pch = 16,
       col = cdr_colors[as.character(dc$CDR)],
       xlab = "Age (years)",
       ylab = "MMSE",
       main = "Longitudinal trajectories of MMSE per subject by CDR group",
       key = list(text = list(c("CDR 0", "CDR 0.5", "CDR 1")),
                  points = list(pch = 16, col = c("lightblue", "violet", "darkblue")),
                  columns = 3))
```

***

```{r}
MMSE_traj
```

***

```{r}
dc_copy <- dc

subject_slopes <- dc_copy %>%
  group_by(Subject.ID, CDR) %>%
  filter(n() > 1) %>%  # need at least two points per subject
  summarise(
    slope = {
      fit <- lm(MMSE ~ Age, data = cur_data())
      coef(fit)["Age"]
    },
    .groups = "drop"
  )

avg_slopes_by_CDR_old <- subject_slopes %>%
  group_by(CDR) %>%
  summarise(
    avg_slope = mean(slope, na.rm = TRUE),
    n = n()
  )
```

***

```{r}
dc$fit_vals5 <- fitted(fit5)

dc_slope <- dc %>%
  select(Subject.ID, Age, fit_vals5, CDR) %>%
  group_by(Subject.ID, CDR) %>%
  arrange(Age, .by_group = TRUE) %>%
  summarise(
    slope = if (n() >= 2) coef(lm(fit_vals5 ~ Age))[2] else NA_real_,
    .groups = "drop"
  )

avg_slopes_by_CDR_5 <- dc_slope %>%
  group_by(CDR) %>%
  summarise(
    avg_slope = mean(slope, na.rm = TRUE),
    n = n()
  )
```

***

```{r}
dc$fit_vals6 <- fitted(fit6)

dc_slope <- dc %>%
  select(Subject.ID, Age, fit_vals6, CDR) %>%
  group_by(Subject.ID, CDR) %>%
  arrange(Age, .by_group = TRUE) %>%
  summarise(
    slope = if (n() >= 2) coef(lm(fit_vals6 ~ Age))[2] else NA_real_,
    .groups = "drop"
  )

avg_slopes_by_CDR_6 <- dc_slope %>%
  group_by(CDR) %>%
  summarise(
    avg_slope = mean(slope, na.rm = TRUE),
    n = n()
  )
```

***

```{r}
dc$fit_spline <- fitted(fit_spline)

dc_slope <- dc %>%
  select(Subject.ID, Age, fit_spline, CDR) %>%
  group_by(Subject.ID, CDR) %>%
  arrange(Age, .by_group = TRUE) %>%
  summarise(
    slope = if (n() >= 2) coef(lm(fit_spline ~ Age))[2] else NA_real_,
    .groups = "drop"
  )

avg_slopes_by_CDR <- dc_slope %>%
  group_by(CDR) %>%
  summarise(
    avg_slope = mean(slope, na.rm = TRUE),
    n = n()
  )
```

***

```{r}
print(avg_slopes_by_CDR_old)
print(avg_slopes_by_CDR_5)
```

***

```{r}
print(avg_slopes_by_CDR_6)
print(avg_slopes_by_CDR)
```

***

- Model with only linear terms is closest for CDR = 1
- Spline model is closest for CDR = 0, CDR = 0.5
- Incorporating non-linearity seems to steepen the mean decline for each CDR group

## References

- Marcus, Fotenos, et. al., (2010). *Open Access Series of Imaging Studies: Longitudinal MRI Data in Nondemented and Demented Older Adults*. MIT Press.

- Marcus, Wang, et al., (2007). *Open Access Series of Imaging Studies (OASIS): cross-sectional MRI data in young, middle aged, nondemented, and demented older adults*. MIT Press.

- Boysen, J., (2017). *MRI and Alzheimers*. Kaggle. Retrieved from https://www.kaggle.com/datasets/jboysen/mri-and-alzheimers

- Morris, J. C., (1993). *The Clinical Dementia Rating (CDR) : Current version and scoring rules*. Wolters Kluwer.

- Folstein, Folstein & McHugh, (1975). *“Mini-mental state”: A practical method for grading the cognitive state of patients for the clinician*. Elsevier.

- Rasmussen & Langerman, (2019). *Alzheimer’s Disease – Why We Need Early Diagnosis*. Dove Medical Press.

## 

<div style="text-align: center; font-size: 60px; font-weight: bold;">
Thanks for your attention!
</div>

